A weakly structured stem for human origins in Africa

### 2023-01-08 paper.tex
Minor format and tense edits

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S -- then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, results in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. While $120$ka can be interpreted as the basal human population divergence,
we suggest concpetualizing this event as the time of most recent shared ancestry among sampled populations.
"Population divergence" in population genetics is a complex concept because of the co-estimation of divergence and migration;
many methods make differing assumptions regarding the degree of a clean split between ancestral groups vs. isolation-with-migration 
(Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in estimates of 
divergence time in the literature (e.g. \citep{Pagani2015-pz }).

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}. Following these merger events, the stems subsequently fracture into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons). We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. Complexity of
thes events warrant the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.

>>>>

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. While $120$ka can be interpreted as the basal human population divergence,
we suggest concpetualizing this event as the time of most recent
shared ancestry among sampled populations.
``Population divergence'' in population genetics is a complex concept because 
of the co-estimation of divergence time and subsequent migration;
methods make differing assumptions regarding the degree of a ``clean'' split 
between descendant groups vs. isolation-with-migration 
(Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of 
divergence time in the literature (e.g.\citep{Pagani2015-pz}).

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons). We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. Complexity of
thes events warrant the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.
